---
title: "Fetch VMM rattenapp data"
author:
- Damiano Oldoni
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

Install required libraries (if not yet installed):

```{r install_pkgs}
installed <- rownames(installed.packages())
required <- c("dplyr", "readr", "purrr", "stringr", "here",
              "readxl", "DBI", "RSQLite", "digest", "tidylog")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r load_pkgs, message = FALSE}
library(dplyr)          # To transform data
library(tidylog)        # To provide feedback on dplyr functions
library(readr)          # To read and write tabular text files
library(purrr)          # To work with lists
library(stringr)        # To work with strings
library(here)           # To find files
library(readxl)         # To read Excel files
library(DBI)            # To create and query databases
library(RSQLite)        # To work with SQLite databases
library(digest)         # To generate hashes
```

# Fetch raw data

The source data are provided manually and stored in `data/raw`. If this workflow can be fully automated, replace with querying from VMM endpoints.

## Fetch observations

### Fetch observations up to 2021

Read both plant and animal observations up to the end of 2021 and merge them:

```{r read_obs_old}
message("Fetch observations up to the end of 2021...")
# Plants up to 2021 included
raw_plants_filename_old <- "Dieren en Planten Waarnemingen 2016-2021.xlsx"
# Get all sheets
raw_plants_sheets_old <- readxl::excel_sheets(
  path = here::here("data", "raw", raw_plants_filename_old)
)
# Get all observations from all sheets in one data.frame
observations_plants_old <- purrr::map_dfr(
  raw_plants_sheets_old, ~readxl::read_xlsx(
    path = here::here("data", "raw", raw_plants_filename_old),
    sheet = .)) %>% 
  mutate(across(everything(), as.character))

# Animals up to 2021 included
raw_animals_filename_old <- "Dieren en Planten Waarnemingen T14_59_52.xlsx"
# Get all sheets
raw_animals_sheets_old <- readxl::excel_sheets(
  path = here::here("data", "raw", raw_animals_filename_old)
)
# Get all observations from all sheets in one data.frame
observations_animals_old <- purrr::map_dfr(
  raw_animals_sheets_old, ~readxl::read_xlsx(
    path = here::here("data", "raw", raw_animals_filename_old),
    sheet = .)) %>% 
  mutate(across(everything(), as.character))

# Merge plants and animals observations up to 2021
observations_old <- dplyr::bind_rows(observations_plants_old,
                                     observations_animals_old)
message("DONE")
```

### Fetch observations from 2022

```{r read_obs_from_2022}
message("Fetch observations from begin 2022...")
raw_obs_filename <- "Dieren en Planten Waarnemingen from 2022-01-01.xlsx"
# Get all sheets
raw_obs_sheets <- readxl::excel_sheets(
  path = here::here("data", "raw", raw_obs_filename)
)
# Get all observations from all sheets in one data.frame
observations <- purrr::map_dfr(
  raw_obs_sheets, ~readxl::read_xlsx(
    path = here::here("data", "raw", raw_obs_filename),
    sheet = .)) %>% 
  mutate(across(everything(), as.character))
message("DONE")
```

Merge all observations:

```{r merge_all_obs}
message("Merge all observations...")
observations <- dplyr::bind_rows(observations_old, observations)
message("DONE")
```

## Fetch catches

### Fetch catches of 2021

Get catches from 2021: 

```{r fetch_catches_old}
message("Fetch catches up to the end of 2021...")
# catches 2021
catches_2021_filename <- "Vangsten T13_05_14.xlsx"
# Get all sheets
catches_2021_sheets <- readxl::excel_sheets(
  path = here::here("data", "raw", catches_2021_filename)
)
# Get all catches from all sheets in one data.frame
catches_2021 <- purrr::map_dfr(
  catches_2021_sheets, ~readxl::read_xlsx(
    path = here::here("data", "raw", catches_2021_filename),
    sheet = .)) %>% 
  mutate(across(everything(), as.character))
message("DONE")
```

### Fetch catches from 2022

Catches from Jan 2022:

```{r catches_from_2022}
message("Fetch catches from begin 2022...")
# catches from Jan 2022
catches_filename <- "Vangsten from 2022-01-01.xlsx"
# Get all sheets
catches_sheets <- readxl::excel_sheets(
  path = here::here("data", "raw", catches_filename)
)
# Get all catches from all sheets in one data.frame
catches <- purrr::map_dfr(
  catches_sheets, ~readxl::read_xlsx(
    path = here::here("data", "raw", catches_filename),
    sheet = .)) %>% 
  mutate(across(everything(), as.character))
message("DONE")
```


```{r merge_catches}
message("Merge all catches...")
catches <- bind_rows(catches_2021, catches)
message("DONE")
message("Relocate columns based on cols observations...")
catches <- 
  catches %>%
  relocate(
    one_of(names(observations)[names(observations) %in% names(catches)])
)
message("DONE")
```

We remove from `observations` the observations related to catches (same registration ID and species name) as they are duplicates:

```{r remove_obs_catches}
message("Remove observations in catches...")
# same registrationID-species combination
observations <- 
  observations %>%
  anti_join(catches %>% dplyr::select(`Registratie ID`,
                                      `Sporen Waarnemingen Naam`),
            by = c("Registratie ID", "Sporen Waarnemingen Naam"))
message("DONE")
```

## Merge observations and catches

We can now proceed by merging `catches` and  `observations`:

```{r merge_catches_obs}
message("Merge observations and catches...")
occurrences <- dplyr::bind_rows(observations, catches)
message("DONE")
```

## Generate hashes for species names

We generate a hash based on the species name as saved in column `Sporen Waarnemingen Naam`. This is needed to create a unique `occurrenceID` of the form `eventID:hash` where `eventID` is the unique identifier of the event as defined in column `Registratie ID`. As long as the species name doesn't change, the hash and so the `occurrenceID` will be stable:

```{r generate_hashes}
vdigest <- Vectorize(digest)
# Generate hashes
occurrences <-
  occurrences %>% 
  mutate(species_name_hash = vdigest(.data$`Sporen Waarnemingen Naam`, algo = "md5"))
```

### Read data already published on GBIF

Read [LIFE MICA dataset](https://www.gbif.org/dataset/3634aee3-41d5-4aa2-8cb5-875859f62a3a) as downloaded from GBIF (https://doi.org/10.15468/dl.ansxse). These data should not be published twice and therefore should be filtered out.

```{r life_mica_data}
life_mica_obs <-
  readr::read_tsv(
    here::here("data", "external", "0218803-210914110416597_occurrence.txt"),
    col_types = cols(.default = "c")) %>%
  
  # Retrieve original event IDs
  mutate(registration_id = stringr::str_remove_all(
    string = .data$eventID, 
    pattern = "MICA:VMM:EV:")
  )
```

# Save data

We save `occurrences` as an interim csv file:

```{r csv_file}
readr::write_csv(occurrences,
                 file = here::here("data", "interim", "raw_occurrences.csv"))
```
